Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-12 2015-eamt-melsa/2015-eamt-melsa.tex
added in an old table, but went a bit over :(
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58894 72bbbca6-d526-0410-a7d9-f06f51895060

@@

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of source-language words, and $G = \{g_1, g_2, \ldots
g_{|G|}\}$ is a set of possible \emph{lexical-selection paths}. A
lexical-selection path $g = (t_1, t_2, \ldots, t_{|S|})$ is a sequence
of lexical-selection choices of those SL words, where, \(t_i=T_s(s_i)\)
and \(T_s(\cdot)\) \todo{Check change \(T(\cdot)\to T_s(\cdot)\) for consistency with later notation} is the set of all translations for a source
word. This is produced in the first stages of rule-based machine
translation, just after morphological analysis, part-of-speech
tagging, and bilingual dictionary lookup, and before any structural
transfer takes place (we will call this \emph{pre-lexsel}). In our
model, it is after these first stages that lexical selection
(\emph{lexsel}) occurs. After lexical selection, structural transfer
and generation take place; a function $\tau(g_i,S)$ represents the
result of these last stages, which we will call \emph{post-lexsel},
and returns a finished translation of a specific lexical-selection
path $g_i$ of sentence $S$. Figure~\ref{fig:lexselschema} shows
 this process schematically.\todo{Can we try to bring the figure closer?}
\begin{figure*}
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.8cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{1.0cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.8cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}
 \caption{A schema of the training process}
 \label{fig:lexselschema}
\end{figure*}

@@

times. We prune features\todo{features or contexts?} occurring less than a certain number of times 

@@

The
whole translation task evaluation is an \emph{extrinsic} evaluation, which
tests how much the system improves as regards final translation
quality in a real system.

@@

Apertium is a free/open-source rule-based machine translation platform, 
it comprises an engine, a toolbox and data to build rule-based machine
translation systems. Translation is implemented as a pipeline consisting
of the following modules: morphological analysis, morphological disambiguation,
lexical transfer, lexical selection, structural transfer and morphological
generation. The relevant \todo{OK?}  modules of the system for this method are the \emph{lexical transfer}
module which provides for each source language lexical form a set of possible
target language forms and the \emph{structural transfer} and \emph{morphological generation}
modules which for each pair of source language lexical form and translation 
produce a final translation in the target language. \todo{The wording of this last sentence is unclear. Identify pre-lexsel, lexsel and post-lexsel as before}

@@

Evaluation will be performed using four language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is at least \todo{check "at least" here} a parallel corpus available, which
is used for evaluation (see Section \ref{ss:measures}). 
% Table~\ref{table:lang-pair-stats} 
%presents statistics about the development status of each of the language 
%pairs chosen, including the number of bilingual dictionary entries, coverage of
%test corpora and translational ambiguity of the bilingual dictionary. Translation
%ambiguity is provided for both all words, and for only ambiguous open-category 
%words. In the experiments, lexical-selection rules are learnt for the open categories
%of adjective, noun and verb.

@@

\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. 
% The resources reused from other pairs
%  were the English morphological dictionary from the Icelandic and
%  English pair \citep{brandt11}, the Macedonian morphological
%  dictionary and constraint grammar from the Macedonian and Bulgarian
%  pair \citep{rangelov11}, and the SETimes parallel corpus
  The lexical resources for the pair were tuned to the SETimes parallel corpus
  \citep{tyers10}.  
% The work was carried out over a period of eight
%  days, and consisted of creating the bilingual dictionary and
%  transfer rules. The bilingual dictionary was created by tagging both
%  sides of the parallel corpus, word-aligning them with {\sc giza++}
%  \citep{och03a} and extracting the probabilistic lexicon. Entries
%  from this lexicon were checked manually according to frequency and
%  included in the bilingual dictionary of the machine translation
%  system. 
  The most probable entry from automatic word alignment of this 
  corpus using {\sc giza++} \citep{och03a} was checked to ensure that it was 
  an adequate translation, and if so marked as the 
  default.\footnote{A bilingual dictionary in Apertium
    \citep{forcada2011apertium} may contain more than one possible
    target-language translation for a given source-language
    word. Where there is more than one possible translation, the
    person who is writing the dictionary chooses the most general or
    most frequent translation among the set of possible translations
    and marks it as the \emph{linguistic default} translation.}\todo{Doesn't this happen in br--fr too?} As a
  result of attempting to include all possible translations, the
  average number of translations per word is much higher % (see table~\ref{table:lang-pair-stats})
than in other pairs.  
% The transfer rules were written
%  by hand to treat translation problems in the corpus. The version of
%  the software used in this paper is SVN revision
%  41476.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}.
\item[Basque--Spanish:] The development of the Basque to Spanish pair
  is described in \cite{ginesti09}. 
% The bilingual dictionary was taken
%  from the free/open-source Matxin system \citep{mayor2011}. 
  Alternative translations were included in the bilingual
  dictionary.% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs
  over a period of around 3--5 months. The pair 
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with
  more than one translation. The most-frequent translation, as judged
  by the language pair developer, is marked as the \emph{default}
  translation and non-default translations added according to
  frequency.\todo{give SVN rev. for all?}\todo{Default is explained for some LPs only, why?} 
% The version used in this paper is SVN revision
%  41387.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}
\end{description}

@@

\subsubsection{Confidence intervals}
Confidence intervals for both metrics will be calculated through the
\emph{bootstrap resampling} method~\citep{efron94} as described
by~\citet{koehn04}. %\footnote{Broadly speaking, % and \citet{zhang04}
%  this involves iteratively taking a random sample (with replacements)
%  from the test set and computing the score for that sample. The
%  scores are sorted and, presuming we want the 95\% confidence
%  interval, the top- and bottom-2.5\% are removed. The highest and
%  lowest remaining are given as the interval. For more detail, refer
%  to the references provided.} 
In all cases, bootstrap resampling will
be carried out for 1,000 iterations. Where the $p = 0.05$ confidence
intervals overlap, we will also perform paired bootstrap resampling
\citep{koehn04}.

@@

For creating the test corpora, and for providing a source language corpus 
for training, and a target language corpus for scoring, we used four parallel corpora: 
\begin{itemize}
\addtolength{\itemsep}{-0.10in}
\item \textbf{Ofis ar Brezhoneg} (\texttt{OAB}): This parallel corpus
  of Breton and French has been collected specifically for
  lexical-selection experiments from translations produced by
  \emph{Ofis ar Brezhoneg} `The Office of the Breton
  language'. %\footnote{In 2010 the administrative status was changed
%    and it was renamed \emph{Ofis Publik ar Brezhoneg} `Public office
%    of the Breton language'. The corpus was published before this date
%    so we use the original name.}  
%  It contains some parallel data
%  previously described in \cite{tyers09b}, and also some new data. 
The corpus has recently been made available online through {\sc opus}.\footnote{\url{http://opus.lingfil.uu.se}}
\item \textbf{South-East European Times} (\texttt{SETimes}): Described
  in \cite{tyers10}, this corpus is a multilingual corpus of the
  Balkan languages (and English) in the news domain. The Macedonian and English
  portion will be used.\todo{Not in OPUS?}
\item \textbf{Open Data Euskadi} (\texttt{OpenData}): This is a Basque
  and Spanish parallel corpus made from the translation memories of
  the \emph{Herri Arduralaritzaren Euskal Erakundea} `Basque Institute
  of Public
  Administration'.\footnote{\url{http://tinyurl.com/eu-es-tm}}\todo{shortened URL}
\item \textbf{European Parliament Proceedings} (\texttt{EuroParl}):
  Described by \cite{koehn05a}, this is a multilingual corpus of the
  European Union official languages. We are using the English--Spanish
  data from version 7.\todo{URL?}
\end{itemize}

@@

%  \begin{table*}
%  \begin{center}
%   \begin{tabular}{|l|r|r|r|r|r|}
%      \hline 
%      {\bf Pair}      & {\bf Lines} & {\bf SL words} & {\bf TL words} & {\bf Amb. words} & {\bf \% ambig} \\ 
%      \hline 
%       \texttt{br-fr} & 1,000 & 13,854  & 13,878  & 1,163 & 8.39   \\
%      \hline 
%       \texttt{mk-en} & 1,000 & 13,441 & 14,228  & 3,872 & 28.80 \\
%      \hline 
%       \texttt{eu-es} & 1,000 & 7,967 & 11,476  & 1,360  & 17.07 \\
%      \hline 
%       \texttt{en-es} & 1,000 & 19,882 & 20,944  & 1,469  & 7.38 \\
%      \hline 
%   \end{tabular} 
%  \end{center}
%   \caption{Statistics about the test corpora. The columns \textbf{amb.\ words} and \textbf{\%~ambig} gives 
%      the number of word with more than one translation and the percentage of SL words which have 
%      more than one translation respectively.}
%    \label{table:test-corp}
%  \end{table*}

After this process, we selected 1,000 sentence pairs at random for
testing (\texttt{test}), 1,000 for development (\texttt{dev})\footnote{The development corpus  
  was used for checking the value for frequency pruning of features.} and 
left the remainder for training. 
%This is a smaller
%number of sentences than used in typical evaluations (such as the WMT
%series, \cite{callisonburch2012}), but was motivated by the fact that
%the smallest corpus (for Breton--French) after sentence discarding was
%only little over 4,000 sentences long. 
Table~\ref{table:input-corp}
gives statistics about the size of the input corpora, and how many
sentences were left after processing for testing, training and
development. \todo{Why do ambiguity numbers change so much from the paper? Having devel and test in different columns would make the caption smaller and the table would still fit}%Table~\ref{table:test-corp} gives information about the
%test corpora.
\todo{``av. amb'' is a corpus average or a dictionary average?}

@@

\begin{table}
 \begin{center}
    \begin{tabular}{|l|r|r|}
        \hline
        \textbf{Pair} & \textbf{Pruned} & \textbf{\# features} \\\hline
        \hline
        {\bf br-fr}  & $< 5$              & 5,277    \\ 
        \hline
        {\bf mk-en} & $< 7$              & 205,494 \\ 
        \hline
        {\bf eu-es} & $< 7$              & 196,024\\
        \hline
        {\bf en-es} & $< 7$              & 195,605\\
        \hline
    \end{tabular}
 \end{center}
 \caption{Number of features (rules) in each rule set and pruning frequency.}
 \label{table:maxent-features}
\end{table}

>>>>

%%% THIS IS IN A WEIRD PLACE BUT WE WANT IT ON PAGE 3 -FMT
\begin{figure*}
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.8cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{1.0cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.8cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}
 \caption{A schema of the training process}
 \label{fig:lexselschema}
\end{figure*}

@@

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of source-language words, and $G = \{g_1, g_2, \ldots
g_{|G|}\}$ is a set of possible \emph{lexical-selection paths}. A
lexical-selection path $g = (t_1, t_2, \ldots, t_{|S|})$ is a sequence
of lexical-selection choices of those SL words, where, \(t_i=T_s(s_i)\)
and \(T_s(\cdot)\) \todo{Check change \(T(\cdot)\to T_s(\cdot)\) for consistency with later notation} is the set of all translations for a source
word. This is produced in the first stages of rule-based machine
translation, just after morphological analysis, part-of-speech
tagging, and bilingual dictionary lookup, and before any structural
transfer takes place (we will call this \emph{pre-lexsel}). In our
model, it is after these first stages that lexical selection
(\emph{lexsel}) occurs. After lexical selection, structural transfer
and generation take place; a function $\tau(g_i,S)$ represents the
result of these last stages, which we will call \emph{post-lexsel},
and returns a finished translation of a specific lexical-selection
path $g_i$ of sentence $S$. Figure~\ref{fig:lexselschema} shows
 this process schematically. %%%\todo{Can we try to bring the figure closer?}

@@

times. We prune features \todo{features or contexts?} occurring less than a certain number of times 

@@

The
whole translation task evaluation is an \emph{extrinsic} evaluation, which
tests how the system improves as regards final translation
quality in a real system.

@@

Apertium is a free/open-source rule-based machine translation platform, 
it comprises an engine, a toolbox and data to build rule-based machine
translation systems. Translation is implemented as a pipeline consisting
of the following modules: morphological analysis, morphological disambiguation,
lexical transfer, lexical selection, structural transfer and morphological
generation. The relevant modules of the system for this method are the \emph{lexical transfer}
module which provides for each source language lexical form a set of possible
target language forms and the \emph{structural transfer} and \emph{morphological generation}
modules which for each pair of source language lexical form and translation 
produce a final translation in the target language. \todo{The wording of this last sentence is unclear. Identify pre-lexsel, lexsel and post-lexsel as before}

@@

Evaluation will be performed using four language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is a parallel corpus available
for evaluation (see Section \ref{ss:measures}). 
% Table~\ref{table:lang-pair-stats} 
%presents statistics about the development status of each of the language 
%pairs chosen, including the number of bilingual dictionary entries, coverage of
%test corpora and translational ambiguity of the bilingual dictionary. Translation
%ambiguity is provided for both all words, and for only ambiguous open-category 
%words. In the experiments, lexical-selection rules are learnt for the open categories
%of adjective, noun and verb.

@@

\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. 
% The resources reused from other pairs
%  were the English morphological dictionary from the Icelandic and
%  English pair \citep{brandt11}, the Macedonian morphological
%  dictionary and constraint grammar from the Macedonian and Bulgarian
%  pair \citep{rangelov11}, and the SETimes parallel corpus
  The lexical resources for the pair were tuned to the SETimes parallel corpus
  \citep{tyers10}.  
% The work was carried out over a period of eight
%  days, and consisted of creating the bilingual dictionary and
%  transfer rules. The bilingual dictionary was created by tagging both
%  sides of the parallel corpus, word-aligning them with {\sc giza++}
%  \citep{och03a} and extracting the probabilistic lexicon. Entries
%  from this lexicon were checked manually according to frequency and
%  included in the bilingual dictionary of the machine translation
%  system. 
  The most probable entry from automatic word alignment of this 
  corpus using {\sc giza++} \citep{och03a} was checked to ensure that it was 
  an adequate translation, and if so marked as the 
  default.\footnote{A bilingual dictionary in Apertium
    \citep{forcada2011apertium} may contain more than one possible
    target-language translation for a given source-language
    word. Where there is more than one possible translation, the
    person who is writing the dictionary chooses the most general or
    most frequent translation among the set of possible translations
    and marks it as the \emph{linguistic default} translation.} As a
  result of attempting to include all possible translations, the
  average number of translations per word is much higher % (see table~\ref{table:lang-pair-stats})
than in other pairs.  
% The transfer rules were written
%  by hand to treat translation problems in the corpus. 
The version of
  the software used in this paper is SVN revision
  41476.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}
\item[Basque--Spanish:] The development of the Basque to Spanish pair
  is described in \cite{ginesti09}. 
% The bilingual dictionary was taken
%  from the free/open-source Matxin system \citep{mayor2011}. 
  Alternative translations were included in the bilingual
  dictionary.\todo{SVN rev}
% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs
  over a period of around 3--5 months. The pair 
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with
  more than one translation. The most-frequent translation, as judged
  by the language pair developer, is marked as the \emph{default}
  translation and non-default translations added according to
  frequency.\todo{Default is explained for some LPs only, why?} 
 The version used in this paper is SVN revision
  41387.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}
\end{description}

@@

\subsubsection{Confidence intervals}
Confidence intervals for both metrics will be calculated through 
\emph{bootstrap resampling}~\citep{efron94} as described
by~\citet{koehn04}. %\footnote{Broadly speaking, % and \citet{zhang04}
%  this involves iteratively taking a random sample (with replacements)
%  from the test set and computing the score for that sample. The
%  scores are sorted and, presuming we want the 95\% confidence
%  interval, the top- and bottom-2.5\% are removed. The highest and
%  lowest remaining are given as the interval. For more detail, refer
%  to the references provided.} 
In all cases, bootstrap resampling will
be carried out for 1,000 iterations. Where the $p = 0.05$ confidence
intervals overlap, we will also perform paired bootstrap resampling
\citep{koehn04}.

@@

For creating the test corpora, and for providing a source language corpus 
for training, and a target language corpus for scoring, we used four parallel corpora: 
\begin{itemize}
\addtolength{\itemsep}{-0.10in}
\item \textbf{Ofis ar Brezhoneg} (\texttt{OAB}): This parallel corpus
  of Breton and French has been collected specifically for
  lexical-selection experiments from translations produced by
  \emph{Ofis ar Brezhoneg} `The Office of the Breton
  language'. %\footnote{In 2010 the administrative status was changed
%    and it was renamed \emph{Ofis Publik ar Brezhoneg} `Public office
%    of the Breton language'. The corpus was published before this date
%    so we use the original name.}  
%  It contains some parallel data
%  previously described in \cite{tyers09b}, and also some new data. 
The corpus has recently been made available online through {\sc opus}.\footnote{\url{http://opus.lingfil.uu.se}}
\item \textbf{South-East European Times} (\texttt{SETimes}): Described
  in \cite{tyers10}, this corpus is a multilingual corpus of the
  Balkan languages (and English) in the news domain. The Macedonian and English
  part will be used.%\todo{Not in OPUS?}
\item \textbf{Open Data Euskadi} (\texttt{OpenData}): This is a Basque
  and Spanish parallel corpus made from the translation memories of
  the \emph{Herri Arduralaritzaren Euskal Erakundea} `Basque Institute
  of Public
  Administration'.\footnote{\url{http://tinyurl.com/eu-es-tm}}
\item \textbf{European Parliament Proceedings} (\texttt{EuroParl}):
  Described by \cite{koehn05a}, this is a multilingual corpus of the
  European Union official languages. We are using the English--Spanish
  data from version 7.\footnote{\url{http://www.statmt.org/europarl/}}
\end{itemize}

@@

  \begin{table}
  \begin{center}
   \begin{tabular}{|l|r|r|r|r|}
      \hline 
      %{\bf Pair}      & {\bf Lines} & {\bf SL words} & {\bf TL words} & {\bf Amb. words} & {\bf \% ambig} \\ 
      {\bf Pair}      & {\bf SL} & {\bf TL} & {\bf Amb.} & {\bf \% amb.} \\ 
      \hline 
       \texttt{br-fr} & 13,854  & 13,878  & 1,163 & 8.39   \\
      \hline 
       \texttt{mk-en} & 13,441 & 14,228  & 3,872 & 28.80 \\
      \hline 
       \texttt{eu-es} & 7,967 & 11,476  & 1,360  & 17.07 \\
      \hline 
       \texttt{en-es} & 19,882 & 20,944  & 1,469  & 7.38 \\
      \hline 
   \end{tabular} 
  \end{center}
   \caption{Statistics about the test corpora. The columns \textbf{SL} and \textbf{TL} give the number
      of tokens in the source and target languages respectively. The columns \textbf{amb.\ words} and \textbf{\%~ambig} gives 
      the number of word with more than one translation and the percentage of SL words which have 
      more than one translation respectively.}
    \label{table:test-corp}
  \end{table}

After this process, we selected 1,000 sentence pairs at random for
testing (\texttt{test}), 1,000 for development (\texttt{dev})\footnote{The development corpus  
  was used for checking the value for frequency pruning of features.} and 
left the remainder for training. 
%This is a smaller
%number of sentences than used in typical evaluations (such as the WMT
%series, \cite{callisonburch2012}), but was motivated by the fact that
%the smallest corpus (for Breton--French) after sentence discarding was
%only little over 4,000 sentences long. 
Table~\ref{table:input-corp}
gives statistics about the size of the input corpora, and how many
sentences were left after processing for testing, training and
development. \todo{Why do ambiguity numbers change so much from the paper? Having devel and test in different columns would make the caption smaller and the table would still fit}Table~\ref{table:test-corp} gives information about the
test corpora.
\todo{``av. amb'' is a corpus average or a dictionary average?}

@@

\begin{table}
 \begin{center}
    \begin{tabular}{|l|r|r|}
        \hline
        \textbf{Pair} & \textbf{Pruned} & \textbf{\# features} \\
        \hline
        {\tt br-fr}  & $< 5$              & 5,277    \\ 
        \hline
        {\tt mk-en} & $< 7$              & 205,494 \\ 
        \hline
        {\tt eu-es} & $< 7$              & 196,024\\
        \hline
        {\tt en-es} & $< 7$              & 195,605\\
        \hline
    \end{tabular}
 \end{center}
 \caption{Number of features (rules) in each rule set and pruning frequency.}
 \label{table:maxent-features}
\end{table}
